---
title: "Homework3"
author: "Devin Teran"
date: "10/4/2021"
output:
  rmdformats::readthedown:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rmdformats)
library(dplyr)
library(ggplot2)
library(dplyr)
library(kableExtra)
library(gridExtra)
require(ISLR)
require(tree)
library(GGally)
library(mice)
library(psych)
library(visdat)
library(tidyr)
```


```{r import-data}
loan_data <- read.csv('https://raw.githubusercontent.com/devinteran/DATA622/main/Loan_approval.csv')
```

## Exploratory Data Analysis
Our loan data has 614 rows and 13 columns, 8 of which are categorical and 5 are numerical.  The target variable is **Loan_Status**, which can be either **Y (yes)** or **N (no)**.  This let's us know if the applicant's loan is approved. There are 7 variables that have blank values. The one with the most blank values is **Credit_History** with 50 blanks.

```{r view}
loan_data %>% kbl() %>% kable_styling() %>% scroll_box(width = "750px", height = "250px")

summary(loan_data) %>% kbl() %>% kable_styling() %>% scroll_box(width = "750px", height = "250px")

missing <- loan_data %>% mutate_if(is.character, list(~na_if(.,""))) 

missing%>%
  summarise_all(list(~sum(is.na(.)))) %>%
  gather(key="Variable", value="Number_Missing") %>%
  arrange(desc(Number_Missing)) %>% kbl() %>% kable_styling() %>% scroll_box(width = "750px", height = "250px")
```

### **Categorical Variables** 
There are several variables which have blank values "".  These data points may have been intentionally skipped by customers from banks during the data collection process or they may just be missing.  We will handle this later on.  
* **Loan_ID:** unique identifier    
* **Gender:** either Male or Female or blank   
* **Married:** either No or Yes or blank  
* **Dependents:** how many dependents does someone have? 0, 1, 2, 3+ or blank  
* **Education:** Graduate or Not Graduate    
* **Self_Employed:**  No or Yes or blank  
* **Property_Area:**  Urban, Rural or Semiurban      
* **Loan_status:** Y (yes) or N (no)    
* **Credit_History:** does the credit history meet the guidelines? 1 = Yes, 0 = No   
 
```{r gender-tables,include=FALSE}
#### Gender  
#More men submitted loan applications than women.  Men have a slightly higher loan approval status than women but not by much. #A small portion of the applications do not list gender.  

gender_loan_status_count <- table(loan_data$Gender,loan_data$Loan_Status)
gender_loan_status_perct <- gender_loan_status_count
gender_loan_status_perct[1,] <- round(gender_loan_status_perct[1,]/13 * 100, 2)
gender_loan_status_perct[2,] <- round(gender_loan_status_perct[2,]/112 * 100, 2)
gender_loan_status_perct[3,] <- round(gender_loan_status_perct[3,]/489 * 100, 2)

#set column names for gender_loan_status_count
gender_loan_status_count <- data.frame(gender_loan_status_count)
colnames(gender_loan_status_count) <- c('Gender','Loan_Status','Count')

#set column names and row names for gender_loan_status_perct
rownames(gender_loan_status_perct) <-  c("Blank", "Female", "Male")
colnames(gender_loan_status_perct) <-  c("% Applications Not Approved", "% Applications Approved")

loan_data_Gender <- loan_data
loan_data_Gender[loan_data_Gender$Gender == '',] <- "Blank"

t1 <- loan_data_Gender %>% group_by(Gender) %>% tally 
colnames(t1) <- c("Gender","Count Loan Applications")
t2 <- gender_loan_status_perct

knitr::kable(list(t1, t2))
```

```{r gender-histogram,include = FALSE}
ggplot(data=gender_loan_status_count, aes(x=Gender, y=Count, fill=Loan_Status)) + geom_bar(stat="identity",position="dodge")
```

#### Married  
Married applicants have a higher approval rate than non married applicants.  It will be useful to look into if this has any correlation with income.  
```{r married-tables,fig.width=7,fig.height=3}
married_loan_status_count <- table(loan_data$Married,loan_data$Loan_Status)
married_loan_status_perct <- married_loan_status_count
married_loan_status_perct[1,] <- round(married_loan_status_perct[1,]/3 * 100, 2)
married_loan_status_perct[2,] <- round(married_loan_status_perct[2,]/213 * 100, 2)
married_loan_status_perct[3,] <- round(married_loan_status_perct[3,]/398 * 100, 2)

#set column names for married_loan_status_count
married_loan_status_count <- data.frame(married_loan_status_count)
colnames(married_loan_status_count) <- c('Married','Loan_Status','Count')

#set column names and row names for gender_loan_status_perct
rownames(married_loan_status_perct) <-  c("Blank", "Not Married", "Married")
colnames(married_loan_status_perct) <-  c("% Applications Not Approved", "% Applications Approved")

loan_data_Married <- loan_data
loan_data_Married[loan_data_Married$Married == '',] <- "Blank"

t1 <- loan_data_Married %>% group_by(Married) %>% tally 
colnames(t1) <- c("Married","Count Loan Applications")
t2 <- married_loan_status_perct

knitr::kable(list(t1, t2)) 

```

```{r married-histogram,fig.width=7,fig.height=3}
ggplot(data=married_loan_status_count, aes(x=Married, y=Count, fill=Loan_Status)) + geom_bar(stat="identity",position="dodge")
```  
  
#### Dependents    
Applicants with 2 dependents appear to have the highest loan approval rate.  It'd be interesting to see if the income per dependent has any impact on loan approval if we assume having more income makes it more likely to get a loan approved.  
```{r dep-tables,fig.width=7,fig.height=3}
dep_loan_status_count <- table(loan_data$Dependents,loan_data$Loan_Status)
dep_loan_status_perct <- dep_loan_status_count
dep_loan_status_perct[1,] <- round(dep_loan_status_perct[1,]/15 * 100, 2)
dep_loan_status_perct[2,] <- round(dep_loan_status_perct[2,]/345 * 100, 2)
dep_loan_status_perct[3,] <- round(dep_loan_status_perct[3,]/102 * 100, 2)
dep_loan_status_perct[4,] <- round(dep_loan_status_perct[4,]/101 * 100, 2)
dep_loan_status_perct[5,] <- round(dep_loan_status_perct[5,]/51 * 100, 2)

#set column names for dep_loan_status_count
dep_loan_status_count <- data.frame(dep_loan_status_count)
colnames(dep_loan_status_count) <- c('Dependents','Loan_Status','Count')

#set column names and row names for gender_loan_status_perct
rownames(dep_loan_status_perct) <-  c("Blank", "0", "1","2","3+")
colnames(dep_loan_status_perct) <-  c("% Applications Not Approved", "% Applications Approved")

loan_data_Dep <- loan_data
loan_data_Dep[loan_data_Dep$Dependents == '',] <- "Blank"

t1 <- loan_data_Dep %>% group_by(Dependents) %>% tally 
colnames(t1) <- c("Dependents","Count Loan Applications")
t2 <- dep_loan_status_perct

knitr::kable(list(t1, t2))
```

```{r dep-histogram,fig.width=7,fig.height=3}
ggplot(data=dep_loan_status_count, aes(x=Dependents, y=Count, fill=Loan_Status)) + geom_bar(stat="identity",position="dodge")
```
  
#### Education    
Applicants with Graduate education have a higher loan approval rate here.  
```{r edu-tables,fig.width=7,fig.height=3}
edu_loan_status_count <- table(loan_data$Education,loan_data$Loan_Status)
edu_loan_status_perct <- edu_loan_status_count
edu_loan_status_perct[1,] <- round(edu_loan_status_perct[1,]/480 * 100, 2)
edu_loan_status_perct[2,] <- round(edu_loan_status_perct[2,]/134 * 100, 2)

#set column names for edu_loan_status_count
edu_loan_status_count <- data.frame(edu_loan_status_count)
colnames(edu_loan_status_count) <- c('Education','Loan_Status','Count')

#set column names for edu_loan_status_perct
colnames(edu_loan_status_perct) <-  c("% Applications Not Approved", "% Applications Approved")

t1 <- loan_data %>% group_by(Education) %>% tally 
colnames(t1) <- c("Education","Count Loan Applications")
t2 <- edu_loan_status_perct

knitr::kable(list(t1, t2))
```

```{r edu-histogram,fig.width=7,fig.height=3}
ggplot(data=edu_loan_status_count, aes(x=Education, y=Count, fill=Loan_Status)) + geom_bar(stat="identity",position="dodge")
```
  
```{r selfemp-tables,include=FALSE}
#### Self Employed    
#The loan approval rate is almost identical here for self employed versus not self employed.  

selfemp_loan_status_count <- table(loan_data$Self_Employed,loan_data$Loan_Status)
selfemp_loan_status_perct <- selfemp_loan_status_count
selfemp_loan_status_perct[1,] <- round(selfemp_loan_status_perct[1,]/32  * 100, 2)
selfemp_loan_status_perct[2,] <- round(selfemp_loan_status_perct[2,]/500  * 100, 2)
selfemp_loan_status_perct[3,] <- round(selfemp_loan_status_perct[3,]/82  * 100, 2)

#set column names for selfemp_loan_status_count
selfemp_loan_status_count <- data.frame(selfemp_loan_status_count)
colnames(selfemp_loan_status_count) <- c('Self_Employed','Loan_Status','Count')

#set column names and row names for selfemp_loan_status_perct
rownames(selfemp_loan_status_perct) <-  c("Blank", "No", "Yes")
colnames(selfemp_loan_status_perct) <-  c("% Applications Not Approved", "% Applications Approved")

loan_data_selfemp<- loan_data
loan_data_selfemp[loan_data_selfemp$Self_Employed == '',] <- "Blank"
t1 <- loan_data_selfemp %>% group_by(Self_Employed) %>% tally 
colnames(t1) <- c("Self_Employed","Count Loan Applications")
t2 <- selfemp_loan_status_perct

knitr::kable(list(t1, t2))
```

```{r self-employed-histogram,include=FALSE}
ggplot(data=selfemp_loan_status_count, aes(x=Self_Employed, y=Count, fill=Loan_Status)) + geom_bar(stat="identity",position="dodge")
```

#### Property Area    
Semiurban applicants have the highest approval loan rating over rural and urban.   
```{r proparea-tables,fig.width=7,fig.height=3}
proparea_loan_status_count <- table(loan_data$Property_Area,loan_data$Loan_Status)
proparea_loan_status_perct <- proparea_loan_status_count
proparea_loan_status_perct[1,] <- round(proparea_loan_status_perct[1,]/179  * 100, 2)
proparea_loan_status_perct[2,] <- round(proparea_loan_status_perct[2,]/233  * 100, 2)
proparea_loan_status_perct[3,] <- round(proparea_loan_status_perct[3,]/202  * 100, 2)

#set column names for proparea_loan_status_count
proparea_loan_status_count <- data.frame(proparea_loan_status_count)
colnames(proparea_loan_status_count) <- c('Property_Area','Loan_Status','Count')

#set column names for proparea_loan_status_perct
colnames(proparea_loan_status_perct) <-  c("% Applications Not Approved", "% Applications Approved")

t1 <- loan_data %>% group_by(Property_Area) %>% tally 
colnames(t1) <- c("Property_Area","Count Loan Applications")
t2 <- proparea_loan_status_perct

knitr::kable(list(t1, t2))
```

```{r property-area-histogram,fig.width=7,fig.height=3}
ggplot(data=proparea_loan_status_count, aes(x=Property_Area, y=Count, fill=Loan_Status)) + geom_bar(stat="identity",position="dodge")
```

#### Credit History  
Having an a credit history that meets the guidelines appears to be extremely important in whether the loan status is approved or not.
```{r credit-history-tables,fig.width=7,fig.height=3}
credhist_loan_status_count <- table(loan_data$Credit_History,loan_data$Loan_Status)
credhist_loan_status_perct <- credhist_loan_status_count
credhist_loan_status_perct[1,] <- round(credhist_loan_status_perct[1,]/89  * 100, 2)
credhist_loan_status_perct[2,] <- round(credhist_loan_status_perct[2,]/475  * 100, 2)

#set column names for credhist_loan_status_count
credhist_loan_status_count <- data.frame(credhist_loan_status_count)
colnames(credhist_loan_status_count) <- c('Credit_History','Loan_Status','Count')

#set column names for credhist_loan_status_perct
colnames(credhist_loan_status_perct) <-  c("% Applications Not Approved", "% Applications Approved")

t1 <- loan_data %>% group_by(Credit_History) %>% tally 
colnames(t1) <- c("Credit_History","Count Loan Applications")
t2 <- credhist_loan_status_perct

knitr::kable(list(t1, t2))
```

```{r credit_history-histogram,fig.width=7,fig.height=3}
ggplot(data=credhist_loan_status_count, aes(x=Credit_History, y=Count, fill=Loan_Status)) + geom_bar(stat="identity",position="dodge")
```

### **Numerical Variables**  
* **ApplicantIncome:**  how much money does the applicant make?    
* **CoapplicantIncome:** how much money does the coapplicant make?  if there is no coapplicant this is 0.      
* **LoanAmount:** how much is the loan worth in thousands?   
* **Loan_Amount_Term:** how many months is the loan?    

Now let's use the **pairs.panels** function to see a lot of important information related to our numeric data:  
  
* **Applicant income** and **loan_amount** are strongly correlated   
* The most common **Loan_Amount_Term** is 360 months  

```{r summary}
numeric_loan_data <- select(loan_data,ApplicantIncome,CoapplicantIncome,LoanAmount,Loan_Amount_Term)
pairs.panels(numeric_loan_data, 
             method = "pearson", # correlation method
             hist.col = "#00AFBB",
             density = TRUE,  # show density plots
             ellipses = TRUE # show correlation ellipses
             )
```

#### Inspecting ApplicantIncome and Loan Income
Here we can see that the **ApplicantIncome** does not have a huge effect on whether the Loan_Status was approved (Y) or not.  The average ApplicantIncome is about the same for both groups is similar.  There are a fewer more outliers of high incomes in the group where the loan status was approved.  
```{r plot-income, warning = FALSE,fig.width=7,fig.height=6}
approved <- loan_data[loan_data$Loan_Status == 'Y',]
denied <- loan_data[loan_data$Loan_Status == 'N',]

a <- ggplot(loan_data,aes(x=ApplicantIncome,color=Loan_Status))  + geom_boxplot()
b <- ggplot(approved,aes(x=ApplicantIncome,y=LoanAmount,color=Loan_Status)) + geom_point(color='blue') + xlab('Approved Applicant Income') + scale_x_continuous(limits = c(0, 25000)) + scale_y_continuous(limits = c(0, 650))
grid.arrange(a,b,nrow=2)#,nrow=2,ncol=2,layout_matrix=c(1,1,2,3)) 
```

```{r plot-income-1,warning = FALSE,fig.width=7,fig.height=3}
c <- ggplot(denied,aes(x=ApplicantIncome,y=LoanAmount,color=Loan_Status)) + geom_point(color='red') + xlab('Denied Applicant Income') + scale_x_continuous(limits = c(0, 25000)) + scale_y_continuous(limits = c(0, 650))

grid.arrange(c)

```

In addition, upon investigating the sum of **ApplicantIncome** and **CoapplicantIncome**, we observe that it does not appear to have much prediction power with **Loan_Status**.

```{r plot-income-combined,warning = FALSE,fig.width=7,fig.height=3}
ggplot(data = loan_data, aes(x = Loan_Status, y = ApplicantIncome+CoapplicantIncome, fill=Loan_Status)) +
  geom_boxplot() +
  coord_flip()
```

#### LoanAmount Per ApplicantIncome  
Now let's see if the rate of the **LoanAmount** divided by **ApplicantIncome** has any prediction power when trying to deteremine if a Loan_Status will be approved or not.  This would indicate that perhaps someone who is requesting a LoanAmount 5 times their income, they might not be approved but if they requested 3 times their income they could get approved.    
  
Looking at the boxplots below, the average LoanAmtPerSalary is roughly the same for approved and not approved applications so this disbunks this theory.  This variable might prove helpful in our modeling so we will keep it.  
```{r plot-loan-per-income,fig.width=7,fig.height=3,warnings=FALSE}
loan_data$LoanAmtPerSalary <- loan_data$LoanAmount*100000/loan_data$ApplicantIncome
ggplot(loan_data,aes(x=LoanAmtPerSalary,color=Loan_Status)) + geom_boxplot() + scale_x_continuous(limits = c(0, 30000))


```




## Missing Data
As we saw in our data exploration, many of the categorical values have been left blank.  We are going to interpret that this was an intentional effort and may ellude to a pattern within those loan applicants.  We will rename these variables as 'Blank'.  
```{r missing-blanks,class.source = 'fold-show'}
clean_loan_data <- loan_data
clean_loan_data[clean_loan_data$Gender == "",]$Gender <- "Blank"
clean_loan_data[clean_loan_data$Married == "",]$Married <- "Blank"
clean_loan_data[clean_loan_data$Dependents == "",]$Dependents <- "Blank"
clean_loan_data[clean_loan_data$Self_Employed == "",]$Self_Employed <- "Blank"

```
  
In terms is other missing data:  
* 14 rows are missing for **loan amount term**  
* 50 rows are missing for **credit history**  
* 14 rows are missing for **loan amount**  
* 22 rows are missing for **loan amount per salary**
  
Since credit history is a categorical value and fewer than 50 rows are missing it's better to delete these data points rather than to try to interpret a value for them.  For loan amount term and loan amount we will use the **mice** package to impute a value where it is missing.  
```{r missing-counts,include=TRUE}
missing_loan_amt_term <- sum(is.na(clean_loan_data$Loan_Amount_Term))
missing_credit_history<- sum(is.na(clean_loan_data$Credit_History))
missing_loan_amt <- sum(is.na(clean_loan_data$Loan_Amount))
missing_loan_amt_per_salary <- sum(is.na(clean_loan_data$LoanAmtPerSalary))

vis_dat(clean_loan_data)
```

```{r impute-missing-numeric-data,class.source = 'fold-show',warnings=FALSE}
clean_loan_data <- clean_loan_data[!is.na(clean_loan_data$Credit_History),]
clean_loan_data <- complete(mice(clean_loan_data,m=5,meth='pmm',print=FALSE))

```


## Data Setup for Modeling
In order to run our models later, we need to make sure that all of our categorical variables are stored as factors.  If we did not, the models will throw errors later in our analysis.  
  
### Factors  
```{r to-factor}
clean_loan_data$Loan_Status = as.factor(clean_loan_data$Loan_Status)
clean_loan_data$Gender = as.factor(clean_loan_data$Gender)
clean_loan_data$Married = as.factor(clean_loan_data$Married)
clean_loan_data$Dependents = as.factor(clean_loan_data$Dependents)
clean_loan_data$Education = as.factor(clean_loan_data$Education)
clean_loan_data$Self_Employed = as.factor(clean_loan_data$Self_Employed)
clean_loan_data$Property_Area = as.factor(clean_loan_data$Property_Area)
```

### Splitting Data into Training & Testing
Here we are going to use 80% of our data to train the model and reserve 20% to test the model we pick.  
```{r train-test}
set.seed(1042)
sample_size <- floor(nrow(clean_loan_data)*0.8)
indices <- sample(1:nrow(clean_loan_data),sample_size)
train <- clean_loan_data[c(indices),]
test <- clean_loan_data[-c(indices),]
```

## Decision Trees  
Now we will use a decision tree to see how well it will perform on our data.    
* Our decision tree starts by splitting users based on their Credit_History.  This makes sense based on our exploratory data analysis.  
* Other variables used in the decision tree include LoanAmount, PropertyArea, CoapplicantIncome 
* It appears that after the initial split, the branch on the left after the root, which splits on LoanAmt < 136.5 has the same outcome on both leaves. Both outcomes are N.  This might show signs of overfitting.  We will address this later.  
```{r decision-tree,fig.width=6,fig.height=6}
loan_tree = tree(Loan_Status ~.-Loan_ID, train)
plot(loan_tree)
text(loan_tree)
title(main = "Unpruned Decision Tree")
```

### Decision Tree Performance  
#### Training Data  
Now we will use our model to see how it performs on the training data. We see that the model predicted **Loan_Status** with an accuracy of 83%.  78 instances were incorrectly classified.  
```{r tree-test}
pred_tree_train <- predict(loan_tree,train,type="class")
test_table <- table(pred_tree_train,train$Loan_Status) %>% kbl() %>% kable_styling()
test_table
mean(pred_tree_train == train$Loan_Status)
```

### Cross-validation for better performance
The first version of our model was a full, unpruned tree.  Now we are going to prune it back to get the optimal tree using cross validation.  We have plotted the number of misclassifications with the different trees.  As we can see, the trees with size 3-6 have the fewest misclassifications.  We will choose size 3 to have the fewest misclassifications and the least complex model.  
```{r cv-trees}
set.seed(2311)
cv_trees = cv.tree(loan_tree,FUN = prune.misclass)
cv_trees
plot(cv_trees)
```
Using a size = 4, our decision tree looks like the following:  
```{r pruned-tree,figure.height=8}
loan_tree_pruned = prune.misclass(loan_tree,best=4)
plot(loan_tree_pruned)
text(loan_tree_pruned)
```

### Testing Data  
Now let's see how our pruned performs on our testing data.  The accuracy for our test data was 81%, which was almost the same as our training data.  21 of the total observations were misclassified.  
```{r confusion-matrix}
pred_tree_test  <- predict(loan_tree_pruned,test, type="class")
test_table <- table(pred_tree_test,test$Loan_Status) %>% kbl() %>% kable_styling()
test_table
mean(pred_tree_test == test$Loan_Status)
```

### linear discriminant analysis

LDA does not seem to be a good approach with this data set as the points provided by the available data are not linearly seperable
```{r}
train%>%
  ggplot(aes(x = log(LoanAmount), y= log(ApplicantIncome+CoapplicantIncome), color = Loan_Status)) + geom_point()
```

### lda model

predictions with the LDA model are less accurate than if we just used the binary classifier Credit_History to determine weather or not a loan would be approved
```{r}
lda = lda(Loan_Status ~ ApplicantIncome+CoapplicantIncome+LoanAmount, data = train)

lda
```

### lda test

```{r}
lda.predictions = predict(lda,test)
```


### Resources:  
* https://www.r-bloggers.com/2015/10/imputing-missing-data-with-r-mice-package/  
* https://daviddalpiaz.github.io/r4sl/trees.html
* https://www.youtube.com/watch?v=YPz2J5lHeVM



